---
title: "Final Assignment Analysis: A multi-decade time series of kelp forest community structure at San Nicolas Islands, California"
author: "Elizabeth Martin, Marjorie Mednikova, Riplee Mercer"
date: "December 8, 2022"
output: pdf_document
fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(dplyr)
```

## Introduction
Kelp is a brown macroalgae that is known for the creation of underwater “forests.” These forests create very biodiverse habitats due to their scale and scope. Their air-filled gas bladders create buoyancy of the fronds, which allow the algae to grow up and towards the sun for maximum photosynthetic ability as well as the creation of larger vertical habitats. With these macroalgae growing up to 200 feet tall, kelp has many uses environmentally and economically. Kelp can build large forests, taking up vertical and horizontal space. Due to their vast range, kelp forests are able to decrease ocean currents, which in turn creates a calmer environment for organisms and reduces wave action and erosion along the coastline. Reduced ocean currents allow many species, including juveniles, to reside in the habitat or use it for resources, which creates high productivity in the forests. Kelp forests are also nitrogen and phosphorus sinks, and are able to increase the pH of water through photosynthesis (Dayton, 1985) . 
El Niño events have historically been shown to impact kelp forest communities. In a study done on the California coast in response to the 1982-83 El Niño event (Tegner et al. 1987), researchers found that large scale changes in the Pacific cause small scale changes in the California ocean current, leading to significant annual variability in physical and biological parameters. These can impact major recruitment events at the local scale as well as impacts from higher sea temperatures and salinities. Temperate kelp forest ecosystems are sensitive to the warm waters and by extension, low nutrient concentrations brought on by El Niño events (Ladah and Zerteche-González, 2007). The impacts of these changes cause mature kelp to reduce reproductive output, reduce recruitment and slough tissue. Kelps exposed to these stressors for too long eventually die, and juveniles under these conditions are especially affected (Ladah and Zerteche-González, 2007). However, there is evidence that microscopic juvenile stages of kelp have the ability to tolerate and recover from El Niño events by acting as a seed bank during the time of the event, although their ability to produce reproductive adults is not yet known (Ladah and Zerteche-González, 2007). 
Sea otters also play a highly influential role within kelp forest communities, as they are top predators within these systems. They are known predators of sea urchins and other grazers that eat macroalgae (Ebeling and Laur, 1988). In areas with high sea otter abundance, urchin populations are diminished, while macroalgae populations flourish due to reduced predation from second-level predators. Increased macroalgal biomass results in a larger sheltered environment for various fish and organisms (Markel and Shurin, 2015; Shelton et al., 2018). Without sea otters, populations of kelp’s predators flourish and can reduce the size of kelp forests substantially, potentially to the point of elimination (Ebeling and Laur, 1988). This trophic cascade is well-documented in various kelp forest systems; however, some recent studies have found that abiotic factors including sea surface temperature and nutrient availability may influence the relationship between otter and kelp abundance, warranting further research into these topics especially as global temperatures rise (Shelton et al. 2018).
Southern sea otters (Enhydra lutris nereis) are listed as threatened under the Endangered Species Act and have been the target of several conservation measures to protect them from possible extinction. In 1987, 140 southern sea otters were translocated from central California to San Nicolas Island in an effort to prevent potential contamination from oil spills (Rathbun et al., 2000; U.S. Fish and Wildlife Service, 2022). San Nicolas was chosen for its ideal subtidal habitat quality as well as its remote location. The size of the reintroduced otter population was monitored for many years following the translocation (Rathbun et al., 2000). Furthermore, a monitoring program documenting algal and benthic invertebrate communities was also implemented in San Nicolas Island in 1980, prior to the planned introduction of sea otters (Kenner et al. 2013).
The dataset from the monitoring program at San Nicolas Island spans over three decades and provides a unique opportunity to observe both short- and long-term temporal changes in kelp communities. The trends observed within this dataset provide insight into seasonal and interannual variability within kelp forests, and help to distinguish these fluctuations from responses to climatic or ecological shifts. In this study, we will use the dataset collected by Kenner et al. (2013) to observe short-term (seasonal) and long-term (decadal) patterns in kelp community abundance. We will also use the dataset to observe trends within these communities both before and after two major types of events: the El Niño Southern Oscillation (ENSO) events in 1982-1983 and 1997-1998, and the reintroduction of southern sea otters to the island in 1987. 


## Pulling in Data

```{r data, include=FALSE}
kelp <-read.csv("final_rawdata.csv",stringsAsFactors = TRUE)
unique(kelp$GENUS_SPECIES)
```


## Data Cleaning

```{r data cleaning, include=FALSE}

## Removing Columns
kelp_cleaned <- kelp %>% 
                  select(-SAMPLE_DESC,
                         -BIOMAS)


## Renaming Columns
kelp_cleaned <- kelp_cleaned %>% 
                        rename(Abundance = ABUNDANCE, 
                               Species = `GENUS_SPECIES`,
                               Plot = PLOT,
                               Latitude = LATITUDE,
                               Longitude = LONGITUDE,
                               Day = DAY, 
                               Month = MONTH,
                               Year = YEAR) 

## Adding Date Column
kelp_cleaned['Date'] <- NA
kelp_cleaned$Date<-as.Date(with(kelp_cleaned,paste(Year,Month,Day,sep="-")),"%Y-%m-%d")


## Adding Phylum Column
phylum <- read.csv("phylum.csv", stringsAsFactors = TRUE)
kelp_cleaned <- left_join(kelp_cleaned, phylum, by="Species")



## Organizing Columns
kelp_cleaned = select(kelp_cleaned, Abundance, Phylum, Species, Plot, Latitude, Longitude, Date, Year, Month, Day)
```

## Methods

*Data cleaning and organization*
Using the raw data from the BioTIME website, a .csv file was downloaded and imported into R Studio using the “read.csv” command. The data was cleaned using various packages contained within tidyverse (including dplyr and ggplot2). The first step was removing any unnecessary columns, which included the empty Biomass column and the Sample Description column, which contained repeated information from other columns. The next step taken was renaming all the columns, to make a cleaner looking dataset. A date column was then added by joining the year, month, and day columns in that order to create a “year-month-day” date. A separate dataset was then pulled in which contained all the phylum names for the 150 species identified in the original dataset. This dataset was then merged using a “left join” to add a phylum column to the original data. Finally all the columns were organized into a cohesive order. 

*Temporal patterns: Short term (seasons) vs. Long term (decades)*
First, to determine whether there were strong seasonal patterns among the data for different phyla, we created a “Seasons” column using a custom R function. This function uses an if-else statement to output the season name based on the month of data collection. The “Seasons” column was added to the dataset using a for loop, with the existing “Month” column as the input. The “summarize” function of dplyr was then used to calculate the mean abundance and standard deviation for each phylum, by season. 

We created a bar graph in ggplot2 using geom_col to show mean abundance by season, faceted by phylum. Error bars showing the standard deviation by phylum were added using geom_errorbar. A one-way analysis of variance (ANOVA) was then run on data within two phyla (Echinodermata and Phaeophyta) to determine the influence of season on abundance.  

We repeated the above procedure to create a “Decade” column through a second custom function and for loop, with the existing “Year” column as the input. We summarized abundance and standard deviation for each phylum, by year. We then created a bar graph showing mean abundance and standard deviation by decade, faceted by phylum. A one-way ANOVA was then conducted assessing the influence of decade on echinoderm abundance. 

*Evaluating responses to sea otter reintroduction and El Nino events*
To determine the impact of the sea otter reintroduction on the echinoderms, including sea urchins, at San Nicolas Island, we plotted the average abundance of echinoderms over time using ggplot2. The “summarize” function of dplyr was used to calculate the mean abundance of echinoderms for each year. A vertical line was added at the position of 1987 along the x-axis to represent the sea otter reintroduction. Shading was added to indicate the 1983 and 1998 El Nino events. 

To visualize the changes within individual echinoderm species over time, we used the “filter” and “summarize” functions again to calculate the mean abundance of each species by year. We then plotted mean abundance over time using ggplot2, using a different color for each decade. 

Next, to assess the response of algal communities to the otter reintroduction, we used dplyr to create a subset of the dataset incorporating the following phyla: Rhodophyta (red algae), Phaeophyta (brown algae), and Chlorophyta (green algae). We then used this subset to graph the mean abundance of each algal phylum over time. We selected by columns abundance, year, date, species and phylum. We then filtered by phylum, grouped by year and phylum, and summarized by the mean abundance. Using the ggplot2 package, we made the x axes=year, y axes=mean abundance, and made each phylum a different color. We then added a line in 1987 showing the reintroduction of sea otters and added shading for the El Nino events.

*Relationship between echinoderm and algal abundance*
To model the relationship between algal abundance and echinoderm abundance, we ran a linear regression analysis between the annual mean abundance values for echinoderms and brown algae. The fitted linear model was graphed using the geom_smooth() function within ggplot. 


## All phyla, over time

```{r time, message=FALSE, fig.cap="\\label{fig:time} Mean annual abundance of each phyla observed at San Nicolas Island, from 1980 to 2011."}

annual.means <- kelp_cleaned %>%
  group_by(Year, Phylum) %>%
  filter(Phylum!="Unknown") %>%
  summarize(MeanAbundance=mean(Abundance))

ggplot(data=annual.means, aes(x=Year, y=MeanAbundance, 
                              by=Phylum, color=Phylum)) +
      geom_point() + geom_line() + 
      labs(x="Year", y="Mean Abundance") + facet_wrap(~Phylum,
      scales="free") + theme(legend.position="none")
  
```

Each phylum exhibited different trends in abundance over time (Figure \ref{fig:time}). Overall increasing abundance was noted for Annelida and Cnidaria. Several phyla exhibited large peaks, including Crustacea, Chlorophyta (green algae), Ochrophyta, and Bryozoa. Notably, many phyla demonstrated large increases in mean abundance within the early 2000s. Finally, Phaeophyta (brown algae) demonstrated an overall decreasing trend, with several notable drops prior to 1985 and 2000. 

## Seasonal patterns, by phylum

```{r Seasons, message=FALSE, fig.cap="\\label{fig:season} Mean seasonal abundance of each phyla observed at San Nicolas Island, from 1980 to 2011."}
# Phylum abundance by season

# Create function using if-else statement to determine season:
season <- function(Month){
  if(Month<=3 | Month>11){return("Winter")
    }else if (Month>=4 & Month<=6){return("Spring")
    }else if(Month>6 & Month<=9){return("Summer")
    }else {return("Fall")}}

kelp_cleaned$Season <- vector("character", length=nrow(kelp_cleaned))

for (i in 1:nrow(kelp_cleaned)){
  kelp_cleaned$Season[i]<- season(kelp_cleaned$Month[i])}

# Average abundance by phylum and season: 
phylum_seasons <- kelp_cleaned %>%
  select(Abundance, Year, Date, Species, Phylum, Season) %>% 
  filter(Phylum!="Unknown") %>%
  group_by(Season, Phylum) %>%
  mutate(Season = factor(Season, levels=c("Spring", "Summer", "Fall", "Winter"))) %>%
  summarize(Mean_Abundance = mean(Abundance), SD = sd(Abundance))

# Plotting average seasonal abundance by phylum, with error bars showing standard deviation:
ggplot(data=phylum_seasons, aes(x=Season, y=Mean_Abundance)) + 
  geom_col(aes(fill=Phylum)) + 
  labs(x="Year", y="Mean Seasonal Abundance") +  
  facet_wrap(~Phylum, scales="free") + 
  geom_errorbar(aes(ymin=Mean_Abundance-SD, ymax=Mean_Abundance+SD),width=.1,color="black") +
  theme_bw()+
  theme(legend.position="none", axis.text.x = element_text(size = 7))

# ANOVA for season, for Phaeophyta and Echinodermata

phaeo.aov <- kelp_cleaned %>%
  filter(Phylum=="Phaeophyta")

phaeo.aov <- summary(aov(data=phaeo.aov, Abundance~Season))

echino.aov <- kelp_cleaned %>%
  filter(Phylum=="Echinodermata")

echino.aov <- summary(aov(data=echino.aov, Abundance~Season))
```
Few phyla exhibited seasonal patterns, especially given the high standard deviation within each season (Figure \ref{fig:season}). Both echinoderms and brown algae (Phaeophyta) were observed to have the highest mean abundance within summer months. Season was found to significantly influence brown algal abundance (p=`r phaeo.aov[[1]][[1,"Pr(>F)"]]`) as well as echinoderm abundance (p=`r echino.aov[[1]][[1,"Pr(>F)"]]`).

## Abundance of Echinoderms over time, with lines for otter reintroduction and El Nino ##

```{r Echinodermata before vs after sea otters, message=FALSE, fig.cap="\\label{fig:echino} Mean abundance of echinoderms observed at San Nicolas Island, from 1980 to 2011. Gray shading indicates 1982-1983 and 1997-1998 El Nino events. The dashed vertical line represents the 1987 reintroduction of sea otters."}

## Subset by phylum
echinodermata_year <- kelp_cleaned %>% 
  select(Abundance, Phylum, Species, Year, Date) %>% 
  filter(Phylum == "Echinodermata") %>% 
  group_by(Year, Phylum) %>% 
  summarize(Mean_Abundance = mean(Abundance))

## Plotting
ggplot(data= echinodermata_year, aes(x= Year, y= Mean_Abundance, 
        by= Phylum, color=Phylum)) +
    geom_rect(aes(xmin=1982, xmax=1983, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
    geom_rect(aes(xmin=1997, xmax=1998, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
    geom_point() + geom_line() + geom_vline(xintercept= 1987,
                  color= 'black', linetype="dashed") + 
  labs(x= "Year", y= "Mean Echinoderm Abundance") + 
  theme(legend.position="none")
```
Echinoderm abundance exhibited notable changes that coincided with multiple environmental events (Figure \ref{fig:echino}). Declines were observed during both El Nino events, in 1982-1983 and 1997-1998. Echinoderm populations also declined following the 1987 sea otter reintroduction. Finally, echinoderm abundance increased substantially following both El Nino events, and returned to pre-otter introduction in the early 2000s. Abundance has oscillated since.  

# Abundance of Echinoderms over time, grouped by decade 

```{r Decades, message=FALSE, fig.height=3, fig.cap="\\label{fig:decades} Mean abundance of echinoderms at San Nicolas Island, by decade."}

#Creating a Decades column:

decades <- function(Year){
  if(Year>=1980 & Year <=1989){return("1980s")
  }else if (Year>=1990 & Year <=1999){return("1990s")
  }else if (Year>=2000 & Year <=2009){return("2000s")
  }else {return("2010s")}}
kelp_cleaned$Decades <- vector("character", length = nrow(kelp_cleaned))
for (i in 1:nrow(kelp_cleaned)){
  kelp_cleaned$Decades[i]<-decades(kelp_cleaned$Year[i])
}

#Abundance of Echinoderms in each decade
echinoderm_decades <- kelp_cleaned %>% 
  select(Abundance, Phylum, Species, Year, Decades) %>% 
  filter(Phylum=="Echinodermata") 
  
# ANOVA testing whether decades is a significant factor in Echinoderm abundance variation:
decades.aov <- summary(aov(data=echinoderm_decades, Abundance~Decades))

# Averages and bar graph
echinoderm_decades2 <- kelp_cleaned %>% 
  select(Abundance, Phylum, Species, Year, Decades) %>% 
  filter(Phylum=="Echinodermata") %>%
  group_by(Decades) %>%
  summarize(MeanAbundance=mean(Abundance), SD=sd(Abundance))

ggplot(data=echinoderm_decades2, aes(x= Decades, y=MeanAbundance)) + geom_col() + labs(x="Decade", y="Mean Echinoderm Abundance") + 
  theme_bw() + theme(legend.position="none", 
                      axis.text.x = element_text(size = 7)) +
  geom_errorbar(aes(ymin=MeanAbundance-SD,
                  ymax=MeanAbundance+SD),width=.1,color="black")

```
The differences in echinoderm abundance observed between decades were found to be significant using a one-way ANOVA, with decade as a factor (p=`r decades.aov[[1]][[1,"Pr(>F)"]]`). The decade with lowest echinoderm abundance was the 1990s (Figure \ref{fig:decades}).

# Echinoderm species, by decade #

```{r decade.facet, message=FALSE, fig.cap="\\label{fig:decade.facet} Mean abundance of each echinoderm species observed at San Nicolas Island, by decade."}

##facet by species

echinoderm_decades3 <- kelp_cleaned %>% 
  select(Abundance, Phylum, Species, Year, Decades) %>% 
  filter(Phylum=="Echinodermata") %>%
  group_by(Year, Decades, Species) %>%
  summarize(MeanAbundance=mean(Abundance))

ggplot(data=echinoderm_decades3,
       aes(x= Year, y=MeanAbundance, color= Decades, by=Species))+
  geom_rect(aes(xmin=1982, xmax=1983, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
  geom_rect(aes(xmin=1997, xmax=1998, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
    geom_point() + geom_line()+
  geom_vline(xintercept= 1987, color= 'black', linetype="dashed")+
facet_wrap(~Species, scales="free") 

```
Echinoderm abundance also varied by species throughout the decades included within this dataset. The 1990s (shown in green in Figure \ref{fig:decade.facet}) consistently demonstrated lower echinoderm abundance among species. Strongylocentrotus species (sea urchins) showed different temporal pattens by species, with S. purpuratus showing highest mean abundances within the 2000s, and S. franciscanus showing highest mean abundances within the 1980s, albeit prior to both the 1982 El Nino event and the 1987 otter reintroduction. 

## Mean abundance of kelp (by phyla) over time, with lines for otter introduction and El Nino #

```{r kelp, message=FALSE, fig.cap="\\label{fig:kelp} Mean abundance of each kelp phylum observed at San Nicolas Island, from 1980 to 2011 Gray shading indicates El Nino events; dashed line indicates sea otter reintroduction."}

# Subsetting by algae: 
algae_year <- kelp_cleaned %>%
  select(Abundance, Year, Date, Species, Phylum) %>% 
  filter(Phylum == "Rhodophyta" | Phylum == "Chlorophyta" |
           Phylum == "Phaeophyta") %>% 
  group_by(Year, Phylum) %>%
  summarize(Mean_Abundance = mean(Abundance))

ggplot(data=algae_year, aes(x=Year, y=Mean_Abundance, 
                            color=Phylum, by=Phylum)) + 
  geom_rect(aes(xmin=1982, xmax=1983, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
  geom_rect(aes(xmin=1997, xmax=1998, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
  geom_point() + geom_line() +
  geom_vline(xintercept= 1987, color= 'black', linetype="dashed")+
  scale_color_manual(values = c("mediumseagreen", "chocolate4", 
                                "red3")) +
  labs(x="Year", y="Mean Annual Abundance")
```
Figure \ref{fig:kelp} shows the change in abundance over time for each phylum of algae. No consistent patterns were observed that applied to all algae phyla; however, some trends were noted within groups. For brown algae (Phaeophyta), declines in abundance took place after both El Nino events, while red algae (Phaeophyta) increased in abundance after both. Brown algae also declined after 1987 (the year of otter reintroduction) following an initial increase. Red algae did not exhibit any notable change in abundance post-reintroduction. Green algal abundance remained relatively low throughout all decades monitored, and did not have a notable response to either El Nino events or otter reintroduction.

## Linear regression between annual abundances of brown algae and echinoderms

```{r lm, message=FALSE, fig.cap="\\label{fig:lm} Relationship between mean annual abundance of echinoderms and brown algae, including best fit line."}

# Echinoderms versus algae, by year #
echino.algae <- kelp_cleaned %>%
  select(Abundance, Year, Date, Species, Phylum) %>%
  filter(Phylum == "Phaeophyta" |Phylum =="Echinodermata") %>%
  group_by(Year, Phylum) %>%
  summarize(Mean_Abundance =mean(Abundance))

echino.algae <- echino.algae %>% 
  pivot_wider(names_from = Phylum, values_from = Mean_Abundance)

ggplot(data=echino.algae, aes(x=Echinodermata, y=Phaeophyta))+
  geom_point() + geom_smooth(method="lm") + labs (x="Echinoderms (annual abundance)", y="Brown algae (annual abundance)")

echino.model <- lm(data=echino.algae, Phaeophyta~Echinodermata)
echino.model <- summary(echino.model)
```
A linear regression model run on mean annual abundance of both echinoderms and brown algae (Figure \ref{fig:lm}) indicated that there was no significant correlation between the two species' abundances (p=`r round(echino.model$coefficients[2,4],2)`). The R squared value of `r round(echino.model$r.squared, 2)` was also low, indicating little to no linear relationship between the two phyla.  



## Discussion

*Temporal patterns*
Long-term temporal patterns in abundance varied substantially among the different phyla monitored at San Nicolas Island. Annelids and cnidarians exhibited fairly constant increasing trends, while large fluctuations were observed within bryozoans, crustaceans, tunicates, and echinoderms. A decreasing trend with several peaks was observed among brown algae. Peaks and fluctuations may be representative of expected cyclical patterns in response to predation within kelp communities, or may reflect variation in water temperatures and weather events from year to year. These findings highlight the importance of long-term monitoring within kelp forest communities, as sampling on a shorter timescale may not adequately show the extent of population declines or expected levels of interannual variability. Given the warm climate of San Nicolas Island, the lack of seasonal patterns in abundance for most phyla was expected. However, echinoderms and brown algae did significantly vary in abundance by season, with highest means observed in summer months. This suggests that they may be particularly sensitive to changes in temperature, and reinforces the importance of consistency in annual sampling times.  

*Echinoderm communities*
Patterns in echinoderm abundance were found to vary substantially between decades. These analyses emphasize the impacts each disturbance to the kelp community had on echinoderms, and how long these disturbances impacted echinoderm abundance. Echinoderm abundance was lowest in the 1990s, directly following the reintroduction of otters, echinoderms’ direct predators (Estes, J.A et al.; 2004). The 1980s exhibited the second highest abundance of echinoderms. Echinoderm abundance increased from 1990-2000, and the highest abundance of echinoderms occurred in the early 2010s, over a decade from the reintroduction of sea otters and El Nino events.

Echinoderm populations decreased during both 1982 and 1997 El Nino events. This response can be attributed to storm surge and turbid waters during El Nino periods (Tegner et al., 1991). These declines were followed by rapid increases in echinoderm populations directly following the El Nino event. This increase is hypothesized to be a result of increased growth rates of echinoderms in warmer water temperatures, which causes an increased grazing rate on kelp forests (Tegner et al., 1991). When sea otters were reintroduced in 1987, echinoderm abundance decreased rapidly. These results are hypothesized to be a direct result of the predator-prey relationship between sea otters and echinoderms (Estes, J.A et al.; 2004). Sea otters prey heavily on echinoderms and have a direct negative impact on echinoderm abundance (Estes, J.A et al.; 2004). 

For individual species of echinoderms, patterns of mean abundance over time differed by species. Lower echinoderm abundance was consistently observed among species in the 1990s, a pattern expected especially for urchins (Strongylocentrotus species) from increased predation following the sea otter reintroduction (Estes, J.A et al.; 2004). Higher species abundances were observed overall both prior to and post-sea otter introduction. This may be a direct result of the impact of sea otter reintroduction combined with El Nino events, and how well certain species were able to recover from these events. 

*Algal communities*
Algal growth is dependent on many factors, including seasonal changes, grazers, temperatures, and water nutrients. The data on the algal populations was divided into the three phylums; Chlorophyta, Rhodophyta, and Phaeophyta. Grazers of algae include echinoderms, commonly sea urchins and starfish, which are known to have increased populations in warmer waters. The graph in Figure \ref{fig:kelp} depicted two shaded areas representing the El Nino events. El Nino events result in an increase in water temperatures, which can allow for echinoderm populations to increase. With higher grazer populations, it is theorized algal populations would plummet (Tegner et al., 1991). This may explain the decreases in brown algae observed following both El Nino events. The other marker on the graph is a line at 1987 indicating the sea otter reintroduction. In theory the introduction of sea otters, predators of echinoderms, should have increased the algal count due to lower grazer populations (Estes, J.A et al.; 2004). Brown algae exhibited initially increased after 1987, but declined shortly thereafter. Green and red algae did not change substantially in abundance following the introduction, and therefore their populations may be more heavily reliant on other environmental factors. 

To highlight whether or not there was a relationship between Phaeophyta populations and echinoderms, a linear regression was run. Phaeophyta include species of kelp, which is a very common food among echinoderms, such as sea urchins. Sea urchin populations have been known to have booms in their populations and create urchin barrens (Dayton, 1985), specifically in kelp forests. This regression demonstrated there was no significant relationship between the two phylums mean abundance per year; it also showed there was no linear relationship between the two. This was not the expected negative linear relationship; echinoderms are grazers of brown algae, and higher grazing should lower food source populations. Since various factors influence the growth and vitality of kelp populations, there isn’t always a clear representation of what ‘should’ occur (Dayton, 1985) and likely there are more factors at play. 

*Future research priorities*
Future research on this kelp community would be helpful to build upon this study in the future. Improvements in future data collection should include more consistency in observation techniques, especially naming of species/phylums that are being observed. There were many species that were named differently throughout the data set, or not consistently named to species level. There were also several “unknown'' species or phyla which made analyses of these observations impossible. Additionally, these data were collected based on abundance alone, which in a long term data set can be observed differently depending on who is collecting the data. More precise measurements such as biomass of each observation would give this data set more credibility and decrease the chances of bias in data collection. This data set focuses on major events such as El Nino and sea otter reintroduction. It would have been helpful to be able to look at sea surface temperature in conjunction with El Nino events which are known for increasing water temperatures. Additional monitoring of environmental parameters would decrease speculation as to what aspects of El Nino events impacted this kelp community. 

## References 

Dayton, P. K. (1985). Ecology of kelp communities. Annual review of ecology and systematics, 215-245.

Dornelas M, Antão LH, Moyes F, Bates, AE, Magurran, AE, et al. BioTIME: A database of biodiversity time series for the Anthropocene. Global Ecol Biogeogr. 2018; 27:760 - 786. https://doi.org/10.1111/geb.12729

Ebeling, A. W., & Laur, D. R. (1988). Fish populations in kelp forests without sea otters: effects of severe storm damage and destructive sea urchin grazing. In The community ecology of sea otters (pp. 169-191). Springer, Berlin, Heidelberg.

Estes, J. A., Danner, E. M., Doak, D. F., Konar, B., Springer, A. M., Steinberg, P. D., ... & Williams, T. M. (2004). Complex trophic interactions in kelp forest ecosystems. Bulletin of marine science, 74(3), 621-638.

Kenner, M. C., Estes, J. A., Tinker, M. T., Bodkin, J. L., Cowen, R. K., Harrold, C., ... & Reed, D. C. (2013). A multi-decade time series of kelp forest community structure at San Nicolas Island, California. Ecological Archives E, 94, 244.

Ladah, L. B., & Zertuche-González, J. A. (2007). Survival of microscopic stages of a perennial kelp (Macrocystis pyrifera) from the center and the southern extreme of its range in the Northern Hemisphere after exposure to simulated El Niño stress. Marine Biology, 152(3), 677-686.

Markel, R. W., & Shurin, J. B. (2015). Indirect effects of sea otters on rockfish (Sebastes spp.) in giant kelp forests. Ecology, 96(11), 2877-2890.

Rathbun, G. B., Hatfield, B. B., & Murphey, T. G. (2000). Status of Translocated Sea Otters at San Nicolas Island, California. The Southwestern Naturalist, 45(3), 322–328. https://doi.org/10.2307/3672835

Shelton, A. O., Harvey, C. J., Samhouri, J. F., Andrews, K. S., Feist, B. E., Frick, K. E., ... & Berry, H. D. (2018). From the predictable to the unexpected: kelp forest and benthic invertebrate community dynamics following decades of sea otter expansion. Oecologia, 188(4), 1105-1119.

Tegner, M. J., & Dayton, P. K. (1987). El Niño effects on southern California kelp forest communities. In Advances in ecological research (Vol. 17, pp. 243-279). Academic Press.

Tegner, M. J., & Dayton, P. K. (1991). Sea urchins, El Ninos, and the long term stability of Southern California kelp forest communities. Marine ecology progress series. Oldendorf, 77(1), 49-63.

U.S. Fish & Wildlife Service. (2022). Southern Sea Otter. Retrieved December 5, 2022 from https://www.fws.gov/species/southern-sea-otter-enhydra-lutris-nereis. 
