---
title: "Results and Analysis"
author: "Elizabeth Martin, Marjorie Mednikova, Riplee Mercer"
date: "November 15, 2022"
output: pdf_document
fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(dplyr)
```

```{r data}
kelp <-read.csv("final_rawdata.csv",stringsAsFactors = TRUE)
```

```{r data cleaning}
## Removing Columns
kelp_cleaned <- kelp %>% 
                  select(-SAMPLE_DESC,
                         -BIOMAS)

## Renaming Columns
kelp_cleaned <- kelp_cleaned %>% 
                        rename(Abundance = ABUNDANCE, 
                               Species = `GENUS_SPECIES`,
                               Plot = PLOT,
                               Latitude = LATITUDE,
                               Longitude = LONGITUDE,
                               Day = DAY, 
                               Month = MONTH,
                               Year = YEAR) 

## Adding Date Column
kelp_cleaned['Date'] <- NA
kelp_cleaned$Date<-as.Date(with(kelp_cleaned,paste(Year,Month,Day,sep="-")),"%Y-%m-%d")

## Adding Phylum Column
phylum <- read.csv("phylum.csv", stringsAsFactors = TRUE)
kelp_cleaned <- left_join(kelp_cleaned, phylum, by="Species")

## Organizing Columns
kelp_cleaned = select(kelp_cleaned, Abundance, Phylum, Species, Plot, Latitude, Longitude, Date, Year, Month, Day)
```
## All phyla, over time

```{r time, message=FALSE, fig.cap="\\label{fig:time} Mean annual abundance of each phyla observed at San Nicolas Island, from 1980 to 2011."}

annual.means <- kelp_cleaned %>%
  group_by(Year, Phylum) %>%
  filter(Phylum!="Unknown") %>%
  summarize(MeanAbundance=mean(Abundance))

ggplot(data=annual.means, aes(x=Year, y=MeanAbundance, 
                              by=Phylum, color=Phylum)) +
      geom_point() + geom_line() + 
      labs(x="Year", y="Mean Abundance") + facet_wrap(~Phylum,
      scales="free") + theme(legend.position="none")
  
```

Each phylum exhibited different trends in abundance over time (Figure \ref{fig:time}). Overall increasing abundance was noted for Annelida and Cnidaria. Several phyla exhibited large peaks, including Crustacea, Chlorophyta (green algae), Ochrophyta, and Bryozoa. Notably, many phyla demonstrated large increases in mean abundance within the early 2000s. Finally, Phaeophyta (brown algae) demonstrated an overall decreasing trend, with several notable drops prior to 1985 and 2000. 

### Seasonal patterns, by phylum

```{r Seasons, message=FALSE, fig.cap="\\label{fig:season} Mean seasonal abundance of each phyla observed at San Nicolas Island, from 1980 to 2011."}
# Phylum abundance by season
# Create function using if-else statement to determine season:
season <- function(Month){
  if(Month<=3 | Month>11){return("Winter")
    }else if (Month>=4 & Month<=6){return("Spring")
    }else if(Month>6 & Month<=9){return("Summer")
    }else {return("Fall")}}
kelp_cleaned$Season <- vector("character", length=nrow(kelp_cleaned))
for (i in 1:nrow(kelp_cleaned)){
  kelp_cleaned$Season[i]<- season(kelp_cleaned$Month[i])}
# Average abundance by phylum and season: 
phylum_seasons <- kelp_cleaned %>%
  select(Abundance, Year, Date, Species, Phylum, Season) %>% 
  filter(Phylum!="Unknown") %>%
  group_by(Season, Phylum) %>%
  mutate(Season = factor(Season, levels=c("Spring", "Summer", "Fall", "Winter"))) %>%
  summarize(Mean_Abundance = mean(Abundance), SD = sd(Abundance))
# Plotting average seasonal abundance by phylum, with error bars showing standard deviation:
ggplot(data=phylum_seasons, aes(x=Season, y=Mean_Abundance)) + 
  geom_col(aes(fill=Phylum)) + 
  labs(x="Year", y="Mean Seasonal Abundance") +  
  facet_wrap(~Phylum, scales="free") + 
  geom_errorbar(aes(ymin=Mean_Abundance-SD, ymax=Mean_Abundance+SD),width=.1,color="black") +
  theme_bw()+
  theme(legend.position="none", axis.text.x = element_text(size = 7))
# ANOVA for season, for Phaeophyta and Echinodermata
phaeo.aov <- kelp_cleaned %>%
  filter(Phylum=="Phaeophyta")
phaeo.aov <- summary(aov(data=phaeo.aov, Abundance~Season))
echino.aov <- kelp_cleaned %>%
  filter(Phylum=="Echinodermata")
echino.aov <- summary(aov(data=echino.aov, Abundance~Season))
```
Few phyla exhibited seasonal patterns, especially given the high standard deviation within each season (Figure \ref{fig:season}). Both echinoderms and brown algae (Phaeophyta) were observed to have the highest mean abundance within summer months. Season was found to significantly influence brown algal abundance (p=`r phaeo.aov[[1]][[1,"Pr(>F)"]]`) as well as echinoderm abundance (p=`r echino.aov[[1]][[1,"Pr(>F)"]]`).

### Abundance of Echinoderms over time, with lines for otter reintroduction and El Nino

```{r Echinodermata before vs after sea otters, message=FALSE, fig.cap="\\label{fig:echino} Mean abundance of echinoderms observed at San Nicolas Island, from 1980 to 2011. Gray shading indicates 1982-1983 and 1997-1998 El Nino events. The dashed vertical line represents the 1987 reintroduction of sea otters."}
## Subset by phylum
echinodermata_year <- kelp_cleaned %>% 
  select(Abundance, Phylum, Species, Year, Date) %>% 
  filter(Phylum == "Echinodermata") %>% 
  group_by(Year, Phylum) %>% 
  summarize(Mean_Abundance = mean(Abundance))
## Plotting
ggplot(data= echinodermata_year, aes(x= Year, y= Mean_Abundance, 
        by= Phylum, color=Phylum)) +
    geom_rect(aes(xmin=1982, xmax=1983, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
    geom_rect(aes(xmin=1997, xmax=1998, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
    geom_point() + geom_line() + geom_vline(xintercept= 1987,
                  color= 'black', linetype="dashed") + 
  labs(x= "Year", y= "Mean Echinoderm Abundance") + 
  theme(legend.position="none")
```
Echinoderm abundance exhibited notable changes that coincided with multiple environmental events (Figure \ref{fig:echino}). Declines were observed during both El Nino events, in 1982-1983 and 1997-1998. Echinoderm populations also declined following the 1987 sea otter reintroduction. Finally, echinoderm abundance increased substantially following both El Nino events, and returned to pre-otter introduction in the early 2000s. Abundance has oscillated since.  

### Abundance of Echinoderms over time, grouped by decade 

```{r Decades, message=FALSE, fig.height=3, fig.cap="\\label{fig:decades} Mean abundance of echinoderms at San Nicolas Island, by decade."}
#Creating a Decades column:
decades <- function(Year){
  if(Year>=1980 & Year <=1989){return("1980s")
  }else if (Year>=1990 & Year <=1999){return("1990s")
  }else if (Year>=2000 & Year <=2009){return("2000s")
  }else {return("2010s")}}
kelp_cleaned$Decades <- vector("character", length = nrow(kelp_cleaned))
for (i in 1:nrow(kelp_cleaned)){
  kelp_cleaned$Decades[i]<-decades(kelp_cleaned$Year[i])
}
#Abundance of Echinoderms in each decade
echinoderm_decades <- kelp_cleaned %>% 
  select(Abundance, Phylum, Species, Year, Decades) %>% 
  filter(Phylum=="Echinodermata") 
  
# ANOVA testing whether decades is a significant factor in Echinoderm abundance variation:
decades.aov <- summary(aov(data=echinoderm_decades, Abundance~Decades))
# Averages and bar graph
echinoderm_decades2 <- kelp_cleaned %>% 
  select(Abundance, Phylum, Species, Year, Decades) %>% 
  filter(Phylum=="Echinodermata") %>%
  group_by(Decades) %>%
  summarize(MeanAbundance=mean(Abundance), SD=sd(Abundance))
ggplot(data=echinoderm_decades2, aes(x= Decades, y=MeanAbundance)) + geom_col() + labs(x="Decade", y="Mean Echinoderm Abundance") + 
  theme_bw() + theme(legend.position="none", 
                      axis.text.x = element_text(size = 7)) +
  geom_errorbar(aes(ymin=MeanAbundance-SD,
                  ymax=MeanAbundance+SD),width=.1,color="black")
```
The differences in echinoderm abundance observed between decades were found to be significant using a one-way ANOVA, with decade as a factor (p=`r decades.aov[[1]][[1,"Pr(>F)"]]`). The decade with lowest echinoderm abundance was the 1990s (Figure \ref{fig:decades}).

### Echinoderm species, by decade

```{r decade.facet, message=FALSE, fig.cap="\\label{fig:decade.facet} Mean abundance of each echinoderm species observed at San Nicolas Island, by decade."}
##facet by species
echinoderm_decades3 <- kelp_cleaned %>% 
  select(Abundance, Phylum, Species, Year, Decades) %>% 
  filter(Phylum=="Echinodermata") %>%
  group_by(Year, Decades, Species) %>%
  summarize(MeanAbundance=mean(Abundance))
ggplot(data=echinoderm_decades3,
       aes(x= Year, y=MeanAbundance, color= Decades, by=Species))+
  geom_rect(aes(xmin=1982, xmax=1983, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
  geom_rect(aes(xmin=1997, xmax=1998, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
    geom_point() + geom_line()+
  geom_vline(xintercept= 1987, color= 'black', linetype="dashed")+
facet_wrap(~Species, scales="free") 
```
Echinoderm abundance also varied by species throughout the decades included within this dataset. The 1990s (shown in green in Figure \ref{fig:decade.facet}) consistently demonstrated lower echinoderm abundance among species. Strongylocentrotus species (sea urchins) showed different temporal pattens by species, with S. purpuratus showing highest mean abundances within the 2000s, and S. franciscanus showing highest mean abundances within the 1980s, albeit prior to both the 1982 El Nino event and the 1987 otter reintroduction. 

### Mean abundance of kelp (by phyla) over time, with lines for otter introduction and El Nino

```{r kelp, message=FALSE, fig.cap="\\label{fig:kelp} Mean abundance of each kelp phylum observed at San Nicolas Island, from 1980 to 2011 Gray shading indicates El Nino events; dashed line indicates sea otter reintroduction."}
# Subsetting by algae: 
algae_year <- kelp_cleaned %>%
  select(Abundance, Year, Date, Species, Phylum) %>% 
  filter(Phylum == "Rhodophyta" | Phylum == "Chlorophyta" |
           Phylum == "Phaeophyta") %>% 
  group_by(Year, Phylum) %>%
  summarize(Mean_Abundance = mean(Abundance))
ggplot(data=algae_year, aes(x=Year, y=Mean_Abundance, 
                            color=Phylum, by=Phylum)) + 
  geom_rect(aes(xmin=1982, xmax=1983, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
  geom_rect(aes(xmin=1997, xmax=1998, ymin=-Inf, ymax=Inf),
                  fill="gray", alpha=0.5, color="gray") +
  geom_point() + geom_line() +
  geom_vline(xintercept= 1987, color= 'black', linetype="dashed")+
  scale_color_manual(values = c("mediumseagreen", "chocolate4", 
                                "red3")) +
  labs(x="Year", y="Mean Annual Abundance")
```
Figure \ref{fig:kelp} shows the change in abundance over time for each phylum of algae. No consistent patterns were observed that applied to all algae phyla; however, some trends were noted within groups. For brown algae (Phaeophyta), declines in abundance took place after both El Nino events, while red algae (Phaeophyta) increased in abundance after both. Brown algae also declined after 1987 (the year of otter reintroduction) following an initial increase. Red algae did not exhibit any notable change in abundance post-reintroduction. Green algal abundance remained relatively low throughout all decades monitored, and did not have a notable response to either El Nino events or otter reintroduction.

### Linear regression between annual abundances of brown algae and echinoderms

```{r lm, message=FALSE, fig.cap="\\label{fig:lm} Relationship between mean annual abundance of echinoderms and brown algae, including best fit line."}
# Echinoderms versus algae, by year #
echino.algae <- kelp_cleaned %>%
  select(Abundance, Year, Date, Species, Phylum) %>%
  filter(Phylum == "Phaeophyta" |Phylum =="Echinodermata") %>%
  group_by(Year, Phylum) %>%
  summarize(Mean_Abundance =mean(Abundance))
echino.algae <- echino.algae %>% 
  pivot_wider(names_from = Phylum, values_from = Mean_Abundance)
ggplot(data=echino.algae, aes(x=Echinodermata, y=Phaeophyta))+
  geom_point() + geom_smooth(method="lm") + labs (x="Echinoderms (annual abundance)", y="Brown algae (annual abundance)")
echino.model <- lm(data=echino.algae, Phaeophyta~Echinodermata)
echino.model <- summary(echino.model)
```
A linear regression model run on mean annual abundance of both echinoderms and brown algae (Figure \ref{fig:lm}) indicated that there was no significant correlation between the two species' abundances (p=`r round(echino.model$coefficients[2,4],2)`). The R squared value of `r round(echino.model$r.squared, 2)` was also low, indicating little to no linear relationship between the two phyla.  
