---
title: "Results and Analysis"
author: "Elizabeth Martin, Marjorie Mednikova, Riplee Mercer"
date: "November 15, 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
```

```{r data}
kelp <-read.csv("final_rawdata.csv",stringsAsFactors = TRUE)
```

```{r data cleaning}
## Removing Columns
kelp_cleaned <- kelp %>% 
                  select(-SAMPLE_DESC,
                         -BIOMAS)

## Renaming Columns
kelp_cleaned <- kelp_cleaned %>% 
                        rename(Abundance = ABUNDANCE, 
                               Species = `GENUS_SPECIES`,
                               Plot = PLOT,
                               Latitude = LATITUDE,
                               Longitude = LONGITUDE,
                               Day = DAY, 
                               Month = MONTH,
                               Year = YEAR) 

## Adding Date Column
kelp_cleaned['Date'] <- NA
kelp_cleaned$Date<-as.Date(with(kelp_cleaned,paste(Year,Month,Day,sep="-")),"%Y-%m-%d")

## Adding Phylum Column
phylum <- read.csv("phylum.csv", stringsAsFactors = TRUE)
kelp_cleaned <- left_join(kelp_cleaned, phylum, by="Species")

## Organizing Columns
kelp_cleaned = select(kelp_cleaned, Abundance, Phylum, Species, Plot, Latitude, Longitude, Date, Year, Month, Day)
```



## Abundance of Echinoderms over time, with lines for otter reintroduction and El Nino ##

```{r Echinodermata w/ sea otters and El Nino events}
## Subset by phylum
echinodermata_year <- kelp_cleaned %>% 
  select(Abundance, Phylum, Species, Year, Date) %>% 
  filter(Phylum == "Echinodermata") %>% 
  group_by(Year, Phylum) %>% 
  summarize(Mean_Abundance = mean(Abundance))

## Plotting
ggplot(data= echinodermata_year,
       aes(x= Year, y= Mean_Abundance,
           color= Phylum, by= Phylum)) +
  geom_point() + geom_line() +
  geom_vline(xintercept= 1987, color= 'blue') + 
  geom_vline(xintercept= 1983, color= 'black') + 
  geom_vline(xintercept= 1998, color= 'black') +
  labs(x= "Year", y= "Mean Annual Abundance")
```

This graph is evaluating the population of echinoderms over time, with lines highlighting the reintroduction of sea otters as well as two El Nino events. The black lines represent the El Nino events, while the blue line represents the reintroduction of sea otters to the San Nicolas Islands. Sea otters are known predators of echinoderms, sea urchins and sea stars especially. It can be theorized that once the population of otters is brought back into the community, the echinoderm population will be negatively impacted due to the predator-prey relationship. It can be see at that the echinoderm population does decrease overtime after the year 1987, when the otters were reintroduced. 



# Abundance of sea urchin species specifically

```{r urchins}
urchin_species <- kelp_cleaned %>%
select(Abundance, Year, Date, Species) %>% 
  filter(Species == "Strongylocentrotus franciscanus" | 
           Species =="Strongylocentrotus purpuratus") %>% 
  group_by(Year, Species) %>%
  summarize(Mean_Abundance = mean(Abundance))

## Plotting
ggplot(data=urchin_species,aes(x=Year, y=Mean_Abundance,
                                color=Species, by=Species)) +
  geom_line(aes(color = Species)) +
  geom_vline(xintercept= 1987, color= 'blue')+
  geom_vline(xintercept= 1983, color= 'black') + 
  geom_vline(xintercept= 1998, color= 'black') +
  labs(x= "Year", y= "Mean Annual Abundance")
```

Sea urhchins are within the phylum echinodermata. They are a known grazer of algae, kelp being a large food source for them. El Nino events produce warmer waters, which in turn increased sea urchin recruitment (Dayton, Tegner 1991). These events are also known to decrease kelp populations. With growing urchin populations and declining kelp forest, there becomes an increase grazing demand. *Strongylocentrotus franciscanus* is the red sea urchin and *Strongylocentrotus purpuratus* is the purple sea urchin. Both urchins live in the kelp forests due to the high food source. 



## Mean abundance of algae (by phyla) over time, with lines for otter introduction and El Nino #

```{r kelp}
## Subsetting by algae: 
algae_year <- kelp_cleaned %>%
  select(Abundance, Year, Date, Species, Phylum) %>% 
  filter(Phylum == "Rhodophyta" | Phylum == "Chlorophyta" |
           Phylum == "Phaeophyta") %>% 
  group_by(Year, Phylum) %>%
  summarize(Mean_Abundance = mean(Abundance))

## Plotting
ggplot(data=algae_year, aes(x=Year, y=Mean_Abundance, 
                            color=Phylum, by=Phylum)) + 
  geom_point() + geom_line() +
  geom_vline(xintercept= 1987, color= 'red') + 
  geom_vline(xintercept= 1983, color= 'blue') + 
  geom_vline(xintercept= 1998, color= 'blue') +
  scale_color_manual(values = c("mediumseagreen", "chocolate4", 
                                "red3")) +
  labs(x="Year", y="Mean Annual Abundance")
```

Chlorophyta is green algae, Rhodophyta is red, and Phaeophyta is brown algae. Chlorophyta levels remain pretty consistant over the time period, while Rhodophyta and Phaeophyta not so much. El Nino events are depicted by black lines and the reintroduction of sea otters by the blue line. Kelp species can be found in the Phaeophyta phylum. El Nino events are known to cause a decrease in kelp forest populations. This can be visualized by the drop in the graph after the black lines in 1983 and 1998. Sea otters are known predators of a well-known grazer of kelp, sea urchins, so after the reintroduction it can be expected kelp populations would increase due to the lack of grazing from urchins. This is not really visualized on the graph though. 
